---
title: "Class 11: Genome informatics"
author: "Alana (PID: A16738319)"
format: pdf
editor: visual
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39790098-40000099;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (3).csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different poplation. I picked the GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G 

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Lets now dig into this further. 

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

How many samples do we have? 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt", head=TRUE)
head(expr)
```

```{r}
str(expr)
```


```{r}
nrow(expr)
```

```{r}
sample_sizes<- table(expr$geno)
sample_sizes
```

```{r}
names(expr) <- c("Sample", "Genotype", "Expression")

# Calculate median expression levels for each genotype
median_expressions <- aggregate(Expression ~ Genotype, data = expr, FUN = median)

# Print the results
print(sample_sizes)
print(median_expressions)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Let's make a boxplot. 

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(x=Genotype , y=Expression, fill=Genotype) + geom_boxplot(notch=TRUE)
```
The boxplot displays expression levels for three genotypes (A/A, A/G, and G/G) of a specific SNP.

The A/A genotype shows higher median expression levels of the gene compared to the G/G genotype. This suggests that the presence of the A allele may be associated with increased expression of the ORMDL3 gene.

The spread of expression values (as seen from the interquartile range and the whiskers of the boxplot) is wider in the A/A genotype compared to the G/G genotype, indicating more variability in expression among individuals with the A/A genotype.

The observed differences in the expression levels across the genotypes A/A and G/G indicate that the SNP likely influences ORMDL3 gene expression. Specifically, individuals with the A/A genotype exhibit higher expression levels than those with the G/G genotype. This suggests that the A allele might enhance the gene's expression. Such a finding is significant, particularly if the ORMDL3 gene is implicated in health conditions like asthma, as it could inform genetic research into how variations influence disease mechanisms and outcomes.



