---
title: "Class 8: Breast Cancer Mini Project"
author: "Alana (PID: A16738319)"
format: pdf
editor: visual
---

## About 

In today's lab we will work with fine needle aspiration (FNA) of breast mass data from the Univesity of Wisconsin. 

# Data Import 

```{r}
fna.data="WisconsinCancer.csv"
wisc.df = read.csv(fna.data, row.names=1)
head(wisc.df)
```

## Exploratory data analysis

>The first step of any data analysis, unsupervised or supervised, is to familiarize yourself with the data.Explore the data you created before (wisc.data and diagnosis) to answer the following questions:

> Q1. How many observations/patients/individual samples are in this dataset? 

569 Observations

> Q2. How many of the observations have a malignant diagnosis? 

```{r}
sum(wisc.df$diagnosis == "M")
```

>table function super useful - it shows 357 binign and 212 malignant 

```{r}
table(wisc.df$diagnosis)
```


> Q3. How many Variables/features in the data are suffixed with _mean?

```{r}
ncol(wisc.df)
```


```{r}
colnames(wisc.df)
```


```{r}
inds <- grep("_mean", colnames(wisc.df))
inds
```

```{r}
length(inds)
```

```{r}
grep("_mean", colnames(wisc.df), value=T)
```

## Initial Analysis 

Before analysis I want to take out the expert diagnosis column (a.k.a the answer) from our dataset. 


```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
head(diagnosis)
```
 
 >this will remove that first diagnosis column and print everything following the first diagnosis column 

```{r}
wisc.data <- wisc.df[,-1]
```




## Clustering 

We can try kmeans() clustering first. 

```{r}
km <- kmeans(wisc.data, centers=2)
```


```{r}
table(km$cluster)
```

Cross-table 

```{r}
table(km$cluster, diagnosis)
```

Let's try `hclust()` the key input required for `hclust()` is a distance matrix as produced by the `dist()` function. 


```{r}
hc <- hclust(dist(wisc.data))
```


I can make a tree like fugure 

```{r}
plot(hc)
```



## PCA

Dp we need to scale the data? 

We can look at the sd of each column (original variable)


```{r}
round(apply(wisc.data, 2, sd))
```



Yes we need to scale. We will run `prcomp()` with `scale=TRUE`.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Generate our main PCA plot (score plot, PC1 vs PC2 plot)...

```{r}
library(ggplot2)

res <- as.data.frame(wisc.pr$x)
```

```{r}
ggplot(res) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```
The PCA plot show a separation of Malignant (turquoise) from Benign (red) samples. 


## Combining methods 

Clustering on PCA results 

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward’s criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.

```{r}
d <- dist(wisc.pr$x[,1:3])   
hc <- hclust(d, method="ward.D2")
plot(hc)
```



To get my clustering results/membership vector I need to "cut" the tree with the `cutree()` function. 


```{r}
grps <- cutree(hc, k=2)
```



> Q. How many patients are in each culster? 

```{r}
table(grps)
```


```{r}
plot(res$PC1, res$PC2, col=grps)
```


## Prediction 

We can use our PCA result (model) to do predictions, that is take new unseen data and project it onto our new PC variables. 


```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```



```{r}
plot(res$PC1, res$PC2, col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], labels=c(1,2), col="white")
```

# Summary 

Principle Component Analysis (PCA) is a super useful method for analyzing large datasets. It works by finding new variables (PCs) that capture the most variance from original variables in your dataset. 








