---
title: "Class 18: Pertussis Vaccination"
author: "Alana (PID: A16738319)"
format: pdf
editor: visual
---

## Investigating pertussis cases by year

Pertussis(whooping cough) is a highly contageous lung iinfection that is most deadly for the very young (under 1 year of age)

Let's begin by having a look at pertussis case numbers per year in the united states.

The CDC tracks pertussis case numbers and makes the data availible here: https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

```{r}
cdc <- data.frame(Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

I want a plot of case numbers per year.

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

base <- ggplot(cdc) + 
  aes(x = Year, y = Cases) + 
  geom_point() + 
  geom_line()

base
```

> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) + 
  aes(x = Year, y = Cases) + 
  geom_point() + 
  geom_line() +
  geom_vline(xintercept=1946,col="blue") +
  geom_vline(xintercept=1996, col="red") +
  geom_vline(xintercept=2020, col="grey")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine (indicated by the red line), pertussis cases initially remained low but began to increase again. Possible explanations for this trend include more sensitive PCR-based testing, bacterial evolution allowing escape from vaccine immunity, waning immunity in adolescents who were vaccinated with the aP vaccine as infants, and increased vaccination hesitancy.

## CMI-PB

A systems vaccinalogy project to figure out whatis going on with aP vs wP immune respnse.

The resouce has an API (application programming interface) that returns JSON format data.

Basically "key": "value" pair format.

We will use the jsonlite package to read this data into R.

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject)
```

> How many individuals/subjects are in this dataset?

```{r}
nrow(subject)
```

> Q. How many wP and aP subjects are there?

```{r}
table(subject$infancy_vac)
```

> Q. How many male and female are there in the dataset?

```{r}
table(subject$biological_sex)
```

> Q. What is the breakdown of race and gender in the dataset?

```{r}
table(subject$race, subject$biological_sex)
```

Read other tables from the CMI-PB resource

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer",simplifyVector = T )
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

I need to link or merge (join) these tables to get all the meta data I need about sibjects and specimens in one place. We will use the **dplyr** `join()` functions for this task.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```

Now we can take our new `meta`

```{r}
abdata <- inner_join(ab_titer, meta)
dim(abdata)
```

```{r}
head(abdata)
```

What Ab are measured/recoreded in the `ab_data` table:

```{r}
table(ab_titer$isotype)
```

```{r}
table(ab_titer$antigen)
```

We have our merged dataset with all the needed metadata and antibody mesaurments called `abdata`

```{r}
head(abdata, 2)
```

## Examine IgG Ab Titer Levels

Now using our joined/merged/linked absata dataset filter() for IgG isotype.

```{r}
igg <- abdata %>% filter (isotype == "IgG")
head(igg)
```

```{r}
base <- ggplot(igg) +
  aes(MFI, antigen) +
geom_boxplot()

base
```

```{r}
table(igg$visit)
```

```{r}
base + 
  facet_wrap(vars(visit), nrow=2)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
table(abdata$dataset)
```

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```


```{r}
library(ggplot2)
ggplot(ssrna) +
  aes(x = visit, y = tpm, group = subject_id) +
  geom_point() +
  geom_line(alpha = 0.2)
```

> What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The expression of the IGHG1 gene reaches its maximum level at visit 4, as indicated by the peak in the tpm values at this time point. This suggests that the highest expression of this gene occurs during this visit, with a significant increase compared to other visits.Additionally, there is considerable variability among subjects, with some showing high expression while others remain low, and a secondary peak at visit 5. By visit 6, the expression levels return to the lower baseline observed at the earlier visits.
